Interpretable Neural Networks using EAGGA

### 2025-03-15 paper/main.tex
abstract

\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
%\blindtext
- tabular data still difficult for NNs, where it's still outperformed by other ML model classes
- research suggests NN performance benefits from heavy regularisation
- using EAGGA, we regularise an NN and achieve both comparable performance as XGB on EAGGA as well as interpretability
- for this, we propose a network architecture specifically suited for the EAGGA algorithm
\end{abstract}

\begin{keywords}
  tabular data, multi-objective optimization, interpretability, deep learning
\end{keywords}

\section{Introduction}
Tabular Data
- most common type of data
- still difficult for neural networks
- \citep[p. 7499]{Borisov_2024}
- recent research suggests that strong regularisation is beneficial to NN performance on tab data \citep[8]{NEURIPS2021_c902b497}

- we know regularisation from linear models can come with improvements in interpretabiltiy
  * e.g. LM-LASSO (L1) regularisation as feature selection -> reduces \# features used in model by setting some coeffs to 0 \citep[p. 267]{lasso}
- other forms of regularisation improve performance
  * e.g. LM-Ridge (L2) on multi-collinear data \citep[p. 55]{ridge}  % TODO: mention in "extension" section that L2 is implemented via weight decay
  * e.g. NN dropout, reduces co-adaptation \citep[p. 1]{hinton2012improvingneuralnetworkspreventing}
  * e.g. NN early stopping, reduces overfitting on training data \citep[p. 778f]{early_stopping}

- we want to explore if we can use NN regularisation to tackle both interpretability and improved performance (i.e. ``comparable'' to XGBoost) on tab data
- using EAGGA framework, which proved it can improve performance while keeping (already high) performance of XGBoost on tab data -> see if interpretabiltiy
  improvements translate to NN + performance can also be on par

%\blindmathpaper

@@

% Why Interpretability is desirable?
Explainability of a model's reasoning is in many ways desirable. \citet[pp. 3-4]{Zach2019InterpretabilityOD} gives a range of examples,
amongst which are \textit{gaining trust}, e.g. when doctors rely on medical diagnosis predictions, avoiding \textit{subconcious biases} by
making sure loan approvals are non-discriminatory, or \textit{regulatory} reasons, most notably the EU's ``Right to Explanation'' warranted by the
GDPR \citep[p. 1]{review_NN_interpretability} or for approval of drugs discovered using machine learning models \citep[1B]{survey_NN_interpretability}.
It can further prove helpful for explaining unexpected drops in model performance, which could arise from optimising a model with respect to loss and then judging its
performance on a different metric such as accuracy, a practice known as \textit{model debugging} \citep[1B]{survey_NN_interpretability} or for
\textit{scientific understanding} in domains where only models can make sense of increasingly complex data anymore and learnt knowledge encoded in a model needs to be
made accessible to humans to be used reliably \citep[p. 1]{review_NN_interpretability}.

@@

Another popular EA approach is Differential Evolution. Its populaton is randomly initialised so the entire search space is covered.
In each generation $g$, for each $\boldsymbol\lambda_{i,g}$ with $i=1,...,n$ it creates ``mutant vectors''
$\boldsymbol\nu_{i,g+1}=\boldsymbol\lambda_{r_1,g}+F\cdot(\boldsymbol\lambda_{r_2,g}-\boldsymbol\lambda_{r_3,g})$,
where $r_1,r_2,r_3\in\{1,...,n\}$ are mutually different random indices that also different from $i$ and $F\in[0,2]$ is constant.
For crossover, it then generates a ``trial vector'' $\boldsymbol\upsilon_{i,g+1}=(\upsilon_{i,g+1}^{(1)},\upsilon_{i,g+1}^{(2)},...,\upsilon_{i,g+1}^{(N)})^T$ with
\begin{equation}
  \upsilon_{i,g+1}^{(j)} = \begin{dcases}
    \nu_{i,g+1}^{(j)} & \text{if } u\le\text{CR}\text{ or } j=R \\
    \lambda_{i,g}^{(j)} & \text{else}
  \end{dcases}
\end{equation}
for a random index $R\in\{1,2,...,D\}$, sampled $u\sim U[0,1]$, and crossover constant $\text{CR}\in[0,1]$.
Selection is done by picking the better of $\boldsymbol\upsilon_{i,g+1}$ and $\boldsymbol\lambda_{i,g}$ with respect to fitness. \citep[p. 343]{differential_evolution}

@@

The authors further introduce a special initialisation of the group structures to increase the sample-efficiency of their algorithm compared to random initialisation.

@@

\subsection{Overall Algorithm}
Our NNs use the following hyperparameters initialised as described
\begin{itemize}
  \item Total number of layers $\in\{3, ..., 10\}$, initialised from a truncated Geometric distribution with $\pi=0.5$.
        This includes input and output layers, the number of hidden layers is thus $\in\{1, ..., 8\}$
  \item Number of nodes per hidden layer $\in\{3, ..., 20\}$, initialisation from a truncated Geometric distribution with $\pi=0.5$
  \item Dropout probability $\in[0, 1]$, initialised from a truncated Gamma distribution with shape 2 and scale 0.15
\end{itemize}
For the group structure initialization we also employ feature, interaction, and monotonicity detectors.
\\
The feature detector is as described by \citet[p. 542]{EAGGA}, but instead of fitting ten trees and using the relative number of features as
probability for the truncated Geometric distribution, we use a constant value of 0.5, as the sklearn DecisionTree examination is not straightforward.
Further, in preliminary experiments we found the sampled number of features to occassionally be larger than the number of non-zero values in the
normalised information gain filter. This can happen if the former is equal to the total number of features and one or more features are independent form the
target. In these edge cases we use all features with non-zero filter-values instead.
\\
The interaction detector similarly uses a constant probability of 0.5 to sample the number of interactions to be used instead of the relative number of interactions
from ten decision trees, for the same reasoning as for the feature detector. Also, for the FAST algorithm, when computing the interaction scores we don't
use the residual sum of squares but the mean accuracy, as the datasets all have binary targets and FAST thus fits a logistic instead of a linear regression.
\\
The monotonicty detector is implemented as described by \citet[p. 543]{EAGGA}, using the default decision tree hyperparameters from mlr3 in Python.
\\
Both the models in the interaction and the monotonicity detector (FAST and an ensemble of decision trees, respectively) use 80\% of the data for model estimation and
the remaining 20\% to compute their scores.

@@

Examining the Pareto sets returned from training, which are accessible on our GitHub repository at ~/code/export/*.csv, we can observe a clear trend in the

@@

the group structure initialised by the detectors.
This was also the reason behind not running EAGGA on the philipine and gina datasets on randomly initialised detectors (as the detectors for those datasets

@@

constrained optimisation in pytorch.

>>>>

\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
  Tabular data proves to be difficult for neural networks, where they are still outperformed by other model classes.
  Following recent research, regularisation can improve their performance. Incorporating the EAGGA framework designed to
  enforce explainability, we improve interpretability and achieve performance comparable to that of XGBoost on the
  metrics AUC, number of features, number of pairwise interactions, and number of non-monotone feature effects.
  For this, we propose a sparse, regularised network architecture specifically suited for EAGGA's constraints on
  interaction and monotonicity.
\end{abstract}

\begin{keywords}
  tabular data, multi-objective optimization, interpretability, neural networks
\end{keywords}

\section{Introduction}
Regularisation has been extensively shown to improve performance from linear models to neural networks.
One of the most popular types of regularisation, Ridge or L2-regularisation is still the preferred method for handling
multicollinear data in linear models \citep[p. 55]{ridge}.
Practicioners training neural networks regularly use regularisation methods such as dropout to reduce
co-adaptation \citep[p. 1]{hinton2012improvingneuralnetworkspreventing} or early stopping to avoid overfitting on the training data \citep[p. 778f]{early_stopping}.
\\
Some regularisation can even increase interpretability, such as LASSO, or L1-regularisation for linear models, which reduces the number of
features used to fit the model by setting some coefficients to 0 \citep[p. 267]{lasso}.
\\
Neural networks as a model class have achieved impressive performance on a multitude of problems, yet still struggle with the most common format --
tabular data \citep[p. 7499]{Borisov_2024}.
Recent research, however, suggest that strong regularisation is beneficial to a neural network's performance on tabular data \citep[8]{NEURIPS2021_c902b497}.
\citet{EAGGA} propose a framework to increase the interpretability of XGBoost using regularisation.
In this report, we explore if the same framework can be extended to neural networks to tackle both interpretability and achieve performance on tabular data
that is on par with XGBoost.

@@

% Why Interpretability is desirable?
Explainability of a model's reasoning is in many ways desirable. \citet[pp. 3-4]{Zach2019InterpretabilityOD} gives a range of examples,
amongst which are \textit{gaining trust}, e.g. when doctors rely on medical diagnosis predictions, avoiding \textit{subconcious biases} by
making sure loan approvals are non-discriminatory, or \textit{regulatory} reasons, most notably the EU's ``Right to Explanation'' warranted by the
GDPR \citep[p. 1]{review_NN_interpretability} or for approval of drugs discovered using machine learning models \citep[1B]{survey_NN_interpretability}.
It can further prove helpful for explaining unexpected drops in model performance, which could arise from optimizing a model with respect to loss and then judging its
performance on a different metric such as accuracy, a practice known as \textit{model debugging} \citep[1B]{survey_NN_interpretability} or for
\textit{scientific understanding} in domains where only models can make sense of increasingly complex data anymore and learnt knowledge encoded in a model needs to be
made accessible to humans to be used reliably \citep[p. 1]{review_NN_interpretability}.

@@

Another popular EA approach is Differential Evolution. Its populaton is randomly initialized so the entire search space is covered.
In each generation $g$, for each $\boldsymbol\lambda_{i,g}$ with $i=1,...,n$ it creates ``mutant vectors''
$\boldsymbol\nu_{i,g+1}=\boldsymbol\lambda_{r_1,g}+F\cdot(\boldsymbol\lambda_{r_2,g}-\boldsymbol\lambda_{r_3,g})$,
where $r_1,r_2,r_3\in\{1,...,n\}$ are mutually different random indices that also different from $i$ and $F\in[0,2]$ is constant.
For crossover, it then generates a ``trial vector'' $\boldsymbol\upsilon_{i,g+1}=(\upsilon_{i,g+1}^{(1)},\upsilon_{i,g+1}^{(2)},...,\upsilon_{i,g+1}^{(N)})^T$ with
\begin{equation}
  \upsilon_{i,g+1}^{(j)} = \begin{dcases}
    \nu_{i,g+1}^{(j)} & \text{if } u\le\text{CR}\text{ or } j=R \\
    \lambda_{i,g}^{(j)} & \text{else}
  \end{dcases}
\end{equation}
for a random index $R\in\{1,2,...,D\}$, sampled $u\sim U[0,1]$, and crossover constant $\text{CR}\in[0,1]$.
Selection is done by picking the better of $\boldsymbol\upsilon_{i,g+1}$ and $\boldsymbol\lambda_{i,g}$ with respect to fitness. \citep[p. 343]{differential_evolution}

@@

The authors further introduce a special initialization of the group structures to increase the sample-efficiency of their algorithm compared to random initialization.

@@

\subsection{Overall Algorithm}
Our NNs use the following hyperparameters initialized as described
\begin{itemize}
  \item Total number of layers $\in\{3, ..., 10\}$, initialized from a truncated Geometric distribution with $\pi=0.5$.
        This includes input and output layers, the number of hidden layers is thus $\in\{1, ..., 8\}$
  \item Number of nodes per hidden layer $\in\{3, ..., 20\}$, initialization from a truncated Geometric distribution with $\pi=0.5$
  \item Dropout probability $\in[0, 1]$, initialized from a truncated Gamma distribution with shape 2 and scale 0.15
\end{itemize}
For the group structure initialization we also employ feature, interaction, and monotonicity detectors.
\\
The feature detector is as described by \citet[p. 542]{EAGGA}, but instead of fitting ten trees and using the relative number of features as
probability for the truncated Geometric distribution, we use a constant value of 0.5, as the sklearn DecisionTree examination is not straightforward.
Further, in preliminary experiments we found the sampled number of features to occassionally be larger than the number of non-zero values in the
normalised information gain filter. This can happen if the former is equal to the total number of features and one or more features are independent form the
target. In these edge cases we use all features with non-zero filter-values instead.
\\
The interaction detector similarly uses a constant probability of 0.5 to sample the number of interactions to be used instead of the relative number of interactions
from ten decision trees, for the same reasoning as for the feature detector. Also, for the FAST algorithm, when computing the interaction scores we don't
use the residual sum of squares but the mean accuracy, as the datasets all have binary targets and FAST thus fits a logistic instead of a linear regression.
\\
The monotonicty detector is implemented as described by \citet[p. 543]{EAGGA}, using the default decision tree hyperparameters from mlr3 in Python.
\\
Both the models in the interaction and the monotonicity detector (FAST and an ensemble of decision trees, respectively) use 80\% of the data for model estimation and
the remaining 20\% to compute their scores.

@@

Examining the Pareto sets returned from training, which are accessible on our GitHub repository\footnote{\url{https://github.com/simon29st/ws2425-tab-ml/}}
at ~/code/export/*.csv, we can observe a clear trend in the

@@

the group structure initialized by the detectors.
This was also the reason behind not running EAGGA on the philipine and gina datasets on randomly initialized detectors (as the detectors for those datasets

@@

constrained optimization in pytorch.
